On Optimality of Finite Element Integration

### 2015-10-31 TOMS_Paper/assembly.tex
Intermediate step with potentially useful sentences

To synthesize optimal implementations as in Definition~\ref{def:mln-optimality} we need:
\begin{itemize}
\item a characterization of the space of legal transformations impacting the operation count
\item a cost model to select the optimal point in the transformation space
\end{itemize}
In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of the transformation space. We consider outer product loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.

@@

In this section, we study optimal \textit{sharing elimination}; that is, a transformation that minimizes a loop nest operation count by applying factorization, code motion and (if necessary) common sub-expression elimination. In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies. However, we explain that by restricting ourselves to the class of loops and expressions in Definition~\ref{def:fem-loopnest}, the search space becomes very small.

From Definition~\ref{def:fem-loopnest} descends the fact that expressions can always be reduced to the form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ (the number of ``terms'') depends on the specific form and discretization, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry. We make the following considerations:

\begin{enumerate}
\item Assume $\alpha \neq \beta$; that is, test and trial functions belong to different function spaces. By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$ and then any $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops are exposed. This systematically exposes all code motion opportunities to minimize the operation count in the innermost loop.
\item In the special case $\alpha = \beta$, common sub-expressions arise over different iteration spaces ($ij$ and $ik$). Compared to point 1), by hoisting and capturing these sub-expressions the reduction loop operation count becomes smaller, although the gain tends to be negligible (the operation count in the innermost loop is not affected).
\item If we neglect the pathological cases in which the loops degenerate (i.e., when the trip count becomes 1, which is the case of P0 discontinuous elements), factorizing and hoisting as in points 1) or 2) systematically reduces the operation count. The reduction is proportional to the outer product loop nest extent.
\item Factorizing any of the terms in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) also impacts the operation count. We distinguish two cases:
\begin{enumerate}
\item Test/trial functions are yet to be factorized. This, in general, inhibits (decreases) code motion by producing non-factorizable, non-hoistable sub-expressions. In the best case (e.g., there is a single scalar coefficient pre-multiplying the form) this simply reduces to cases 1) or 2).
\item Within $\sigma_{ei}$ itself; for example, after applying point 1) we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$. Splitting $L_e$-dependent from $L_i$-dependent sub-expressions (through suitable re-factorization) and applying code motion naturally tends to operation count minimization. Similarly to point 3), however, with very low trip counts this strategy may be sub-optimal. For example, consider the following sub-expression, $(a v_i + b w_i) (c v_i + d w_i)$, which is excerpted from a multiplication of the inverse Jacobian matrix for the change of coordinates with the derivative of a vector-valued coefficient. It is true that factorizing $v_i$ and $w_i$ exposes $L_i$-independent sub-expressions, but this will actually increase the operation count when the size of $L_i$ is 1.
\end{enumerate}
\end{enumerate}

@@

\subsection{Pre-evaluation}
\label{sec:pre-evaluation}
Sharing elimination reduces the operation count by applying three operators: factorization, code motion and/or common sub-expression elimination. A fourth operator impacting the operation count is \textit{reduction pre-evaluation}. In this section, we discuss its role and show that there exists a single point in the whole transformation space (i.e., a specific factorization) that makes applying this operator legal.

>>>>

To synthesize optimal implementations as in Definition~\ref{def:mln-optimality} we need:
\begin{itemize}
\item a characterization of the space of legal transformations impacting the operation count
\item a cost model to select the optimal point in the transformation space
\end{itemize}
In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of such transformation space. We consider outer product loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.

@@

In this section, we study optimal \textit{sharing elimination}; that is, a transformation that minimizes a loop nest operation count by applying factorization, code motion and (if necessary) common sub-expression elimination. In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies. However, we show that by restricting ourselves to the class of loops and expressions in Definition~\ref{def:fem-loopnest}, the search space becomes very small.

Finite element integration expressions can be reduced to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ (the number of ``terms'') depends on the problem, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry. Note that this respects Definition~\ref{def:fem-loopnest}. We make the following considerations:

\begin{enumerate}
\item In the special case in which test and trial functions are in the same function space, there exists $w1$ and $w2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$. That is, common sub-expressions arise over different iteration spaces ($L_j$ and $L_k$). By hoisting $\alpha$ or $\beta$ at the level of the closest common loop ($L_i$), the operation count is reduced.
\item Let us now consider the general case in which test and trial functions belong to different function spaces. By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$ and then any $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops are exposed (a simplistic example is showed in Figure~\ref{code:multi_loopnest}). Similarly to point 1), such a factorization aims to minimize the operation count ``locally'' at the level of the outer product loop nest. This transformation is independent of the chosen function spaces, but it may not be optimal in case point 1) is applicable (a common sub-expression elimination opportunity is lost). Two observations follow:
\begin{enumerate}
\item In case both 1) and 2) are applicable, there is no simple way of predicting what transformation provides the largest decrease in operation count; 
\item However, the difference between the two tends to be negligible, since the outer product loop nests eventually perform the same number of operations.
\end{enumerate}
\item If we neglect the pathological cases in which the loops degenerate (i.e., when the trip count becomes 1, the case of P0 discontinuous elements), points 1) and 2) always result in reduced operation count.
\item Factorizing any of the symbols in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) also impacts the operation count. We distinguish two cases:
\begin{enumerate}
\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form can be reduced to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus f_i$. 
\item There are no common symbols among the terms. Factorizing any $f_i \in \sigma_{ei}$ would then

\item The starting point of the factorization is the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$. Since the symbols in $\sigma_{ei}$ are generally not common factors

In this case, the production of non-factorizable, non-hoistable sub-expressions tends to drastically inhibit code motion. In the best case (e.g., there is a single scalar coefficient pre-multiplying the form) this simply reduces to cases 1) or 2). In other words, this is a ``non-natural factorization'', which tries to group potentially non-common factors and does not take advantage of loop dependencies.
\item Point 1) was applied -- for example, we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$ -- and we factorize symbols within $\sigma_{ei}$. Splitting $L_e$-dependent from $L_i$-dependent sub-expressions (through suitable re-factorization) and applying code motion naturally tends to operation count minimization. Similarly to point 3), however, with very low trip counts this strategy may be sub-optimal. For example, consider the following sub-expression, $(a v_i + b w_i) (c v_i + d w_i)$, which is excerpted from a multiplication of the inverse Jacobian matrix for the change of coordinates with the derivative of a vector-valued coefficient. It is true that factorizing $v_i$ and $w_i$ exposes $L_i$-independent sub-expressions, but this may increase the operation count when the size of $L_i$ is 1.
\end{enumerate}
\end{enumerate}

@@

\subsection{Pre-evaluation}
\label{sec:pre-evaluation}
Sharing elimination reduces the operation count by applying three operators: factorization, code motion and/or common sub-expression elimination. A fourth operator impacting the operation count is \textit{reduction pre-evaluation}. In this section, we discuss its role and explain that there exists a single point in the whole transformation space (i.e., a specific factorization) that makes applying this operator legal.
